Evaluating Cloud-Optimized HDF5 for NASA's ICESat-2 Mission

### 2024-10-03 _tex/paper.tex
Built site for gh-pages

As a result of community feedback and ``hack weeks'' organized by NSIDC
and UW eScience Institute in 2023(ICESAT-2 HackWeek, 2023), NSIDC
started the Cloud Optimized Format Investigation (COFI) project to
improve access to HDF5 from the ICESat-2 mission. A spaceborne lidar
that retrieves surface topography of the Earth's ice sheets, land and
oceans (Neumann et al., 2019). Because of its complexity, large size and
importance for cryospheric studies we targeted the ATL03 dataset. ATL03
core data are geolocated photon heights from the ICESat-2 ATLAS
instrument. Each file contains 1003 geophysical variables in 6 data
groups. Although our research was focused on this dataset, most of our
findings are applicable to any dataset stored in HDF5 and NetCDF4.

@@

We tested access times to original and different configurations of
cloud-optimized HDF5 ATL03 files {[}list files tested{]} stored in AWS
S3 buckets in region us-west-2, the region hosting NASA's Earthdata
Cloud archives. Files were accessed using Python tools commonly used by
Earth scientists: h5py and Xarray(Hoyer \& Hamman, 2017). h5py is a
Python wrapper around the HDF5 C API. xarray\footnote{\texttt{h5py} is a
  dependency of Xarray} is a widely used Python package for working with
n-dimensional data. We also tested access times using h5coro, a python
package optimized for reading HDF5 files from S3 buckets and kerchunk, a
tool that creates an efficient lookup table for file chunks to allow
performant partial reads of files.

The test files were originally cloud optimized by ``repacking'' them,
using a relatively new feature in the HDF5 C API called ``paged
aggregation''. Page aggregation does 2 things: it collects file-level
metadata from datasets and stores it on dedicated metadata blocks in the
file; and it forces the library to write both data and metadata using
these fixed-size pages. Aggregation allows client libraries to read file
metadata with only a few requests and uses the page size used in the
aggregation as the minimal request size, overriding the 1 request per
chunk behavior.

@@

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.1589}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.3444}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.1457}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.0993}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.0728}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.0728}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.1060}}@{}}
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
prefix
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
description
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
\% file size increase
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
\textasciitilde km per chunk
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
shape
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
page size
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
avg\_chunk\_size
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
original & original file from ATL03 v6 (1gb and 7gb) & 0 & 1.5km &
(10000,) & N/A & 35kb \\
original-kerchunk & kerchunk sidecar of the original file & N/A & 1.5km
& (10000,) & N/A & 35kb \\
page-only-4mb & paged-aggregated file with 4mb per page &
\textasciitilde1\% & 1.5km & (10000,) & 4MB & 35kb \\
page-only-8mb & paged-aggregated file with 4mb per pag8 &
\textasciitilde1\% & 1.5km & (10000,) & 8MB & 35kb \\
rechunked-4mb & page-aggregated and bigger chunk sizes &
\textasciitilde1\% & 10km & (100000,) & 4MB & 400kb \\
rechunked-8mb & page-aggregated and bigger chunk sizes &
\textasciitilde1\% & 10km & (100000,) & 8MB & 400kb \\
rechunked-8mb-kerchunk & kerchunk sidecar of the last paged-aggregated
file & N/A & 10km & (10000,) & 8MB & 400kb \\
\end{longtable}

@@

@@

\caption{\label{fig-5}Benchmarks show that cloud optimizing ATL03 files
improved access times at least an order of magnitude when used with
aligned I/O patterns, this is telling the library about the cloud
optimization and page size.}

>>>>

As a result of community feedback and ``hack weeks'' organized by NSIDC
and UW eScience Institute in 2023(ICESAT-2 HackWeek, 2023), NSIDC
started the Cloud Optimized Format Investigation (COFI) project to
improve access to HDF5 from the ICESat-2 mission, a spaceborne lidar
that retrieves surface topography of the Earth's ice sheets, land and
oceans (Neumann et al., 2019). Because of its complexity, large size and
importance for cryospheric studies we targeted the ATL03 data product.
The most relevant variable in ATL03 are geolocated photon heights from
the ICESat-2 ATLAS instrument. Each ATL03 file contains 1003 geophysical
variables in 6 data groups. Although our research was focused on this
dataset, most of our findings are applicable to any dataset stored in
HDF5 and NetCDF4.

@@

We tested access times to original and different configurations of
cloud-optimized HDF5
\href{https://its-live-data.s3.amazonaws.com/index.html\#test-space/cloud-experiments/h5cloud/}{ATL03
files} stored in AWS S3 buckets in region us-west-2, the region hosting
NASA's Earthdata Cloud archives. Files were accessed using Python tools
commonly used by Earth scientists: h5py and Xarray(Hoyer \& Hamman,
2017). h5py is a Python wrapper around the HDF5 C API. xarray\footnote{\texttt{h5py}
  is a dependency of Xarray} is a widely used Python package for working
with n-dimensional data. We also tested access times using h5coro, a
python package optimized for reading HDF5 files from S3 buckets and
kerchunk, a tool that creates an efficient lookup table for file chunks
to allow performant partial reads of files.

The test files were originally cloud optimized by ``repacking'' them,
using a relatively new feature in the HDF5 C API called ``paged
aggregation''. Page aggregation does 2 things: fisrt, it collects
file-level metadata from datasets and stores it on dedicated metadata
blocks at the front of the file; second, it forces the library to write
both data and metadata using these fixed-size pages. Aggregation allows
client libraries to read file metadata with only a few requests using
the page size as a fixed request size, overriding the 1 request per
chunk behavior.

@@

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.1589}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.3444}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.1457}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.0993}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.0728}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.0728}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.1060}}@{}}
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
prefix
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
description
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
\% file size increase
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
\textasciitilde km per chunk
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
shape
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
page size
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
avg\_chunk\_size
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
original & original file from ATL03 v6 (1gb and 7gb) & 0 & 1.5km &
(10000,) & N/A & 35kb \\
original-kerchunk & kerchunk sidecar of the original file & N/A & 1.5km
& (10000,) & N/A & 35kb \\
page-only-4mb & paged-aggregated file with 4mb per page &
\textasciitilde1\% & 1.5km & (10000,) & 4MB & 35kb \\
page-only-8mb & paged-aggregated file with 4mb per pag8 &
\textasciitilde1\% & 1.5km & (10000,) & 8MB & 35kb \\
rechunked-4mb & page-aggregated and bigger chunk sizes &
\textasciitilde1\% & 10km & (100000,) & 4MB & 400kb \\
rechunked-8mb & page-aggregated and bigger chunk sizes &
\textasciitilde1\% & 10km & (100000,) & 8MB & 400kb \\
rechunked-8mb-kerchunk & kerchunk sidecar of the last paged-aggregated
file & N/A & 10km & (100000,) & 8MB & 400kb \\
\end{longtable}

@@

\includegraphics{figures/figure-4.png}

\caption{\label{fig-4}shows that using paged aggregation alone is not a
complete solution. This behavior us caused by over-reads of data now
distributed in pages and the internals of HDF5 not knowing how to
optimize the requests. This means that if we cloud optimize alone and
use the same code, in some cases we'll make access to these files even
slower. A very important thing to notice here is that rechunking the
file, in this case using 10X bigger chunks results in a predictable 10X
improvement in access times without any cloud optimization involved.
Having less chunks generates less metadata and bigger requests, in
general is it recommended that chunk sizes should range between 1MB and
10MB{[}Add citation, S3 and HDF5{]} and if we have anough memory and
bandwith even bigger (Pangeo recommends up to 100MB chunks){[}Add
citation.{]}}

@@

\caption{\label{fig-5}shows that performance once the I/O configuration
is aligned with the chunking in the file, access times perform on par
with cloud optimized access patterns like Kerchunk/Zarr. These numbers
are from in-region execution. Out of region is considerable slower for
the non cloud optimized case.}
